---
title: "Sanders Limestone Fine Sampling Analysis"
author: "Limefine Project"
format: html
editor: visual
---

## Introduction

This document contains the analysis for the Sanders limestone fine sampling campaign.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required libraries
# Install packages if needed
required_packages <- c("tidyverse", "ggplot2", "readxl", "patchwork", "corrplot", "Hmisc")
missing_packages <- required_packages[!required_packages %in% installed.packages()[,"Package"]]
if(length(missing_packages) > 0) {
  install.packages(missing_packages)
}

# Load required libraries
library(tidyverse)
library(ggplot2)
library(readxl)
library(patchwork)
library(corrplot)
library(Hmisc)
```

## Data Loading

### Fines data

```{r load-data}
# Load data from input folder

dt_fines <- read_excel("input/Fines_Sample_data.xlsx",
    sheet = "XRF", skip = 1)

```

## Data Exploration

```{r explore}
# Summary statistics
summary(dt_fines)



names(dt_fines) <- c("Mineralresource", "Deposit", "Samplenumber", "LOI950C", "SiO2", "TiO2", "Al2O3",
"Fe2O3", "MnO", "MgO", "CaO", "Na2O", "K2O", "P2O5",
"SO3")


```

## Plots

### Boxplot of XRF results fines (regular y-axis)

```{r boxplot-xrf-prep}

# format data for plotting
# pivot longer, excluding non-numeric columns

#exclude LOI950°C 
dt_fines_long <- dt_fines %>%  select(-c("LOI950C"))

dt_fines_long <- dt_fines_long %>%
  pivot_longer(cols = c("SiO2", "TiO2", "Al2O3",
"Fe2O3", "MnO", "MgO", "CaO", "Na2O", "K2O", "P2O5",
"SO3"),
               names_to = "Substance",
               values_to = "Concentration")

# create boxplot, sort x-axis by median concentration
#plot  MgO, SiO2, CaO separately due to scale issues
#start with these

subs_high <- c("MgO",  "SiO2",  "CaO")
subs_medium <- c("Al2O3", "Fe2O3", "K2O")
subs_low <- c("TiO2", "MnO", "P2O5", "Na2O", "SO3")

#refactor the three plots below into a function


```

```{r boxplot-xrf}

# Create a named vector for chemical formula labels with subscripts
chem_labels <- c(
  "SiO2" = expression(SiO[2]),
  "TiO2" = expression(TiO[2]),
  "Al2O3" = expression(Al[2]*O[3]),
  "Fe2O3" = expression(Fe[2]*O[3]),
  "MnO" = expression(MnO),
  "MgO" = expression(MgO),
  "CaO" = expression(CaO),
  "Na2O" = expression(Na[2]*O),
  "K2O" = expression(K[2]*O),
  "P2O5" = expression(P[2]*O[5]),
  "SO3" = expression(SO[3])
)

boxplot_fines <- function(substances){
  ggplot(dt_fines_long %>% filter(Substance %in% substances),
         aes(x = reorder(Substance, Concentration, FUN = median), y = Concentration)) +
    geom_boxplot(outliers = F,
                 size=0.3) +
    labs(x = "Substance",
         y = "Concentration (%)") +
    scale_x_discrete(labels = chem_labels) +
    theme_bw()
  #  geom_jitter(size = 0.1, alpha = 0.5)
}


a <- boxplot_fines(subs_low)

b <- boxplot_fines(subs_medium)
   
c <- boxplot_fines(subs_high)


# combine plots into a single patchwork object
p <- (a + b + c) + plot_layout(nrow = 1, guides = "collect", axes = "collect")

p_combined <- (a + b + c) + plot_layout(nrow = 1, guides = "collect", axes = "collect")
p_combined  # display combined plot in the document
  #apply theme_bw to all plots


#save to file (png, width=12in, height=6in, res=300)

ggsave("output/boxplot_xrf_fines.png", plot = p_combined, width = 12, height = 6, dpi = 300, scale = 0.6)




```

### Boxplot of XRF results fines (log y-axis)

```{r boxplot-xrf-log}

# create boxplot with log y-axis, sort x-axis by median concentration
a_log <- ggplot(dt_fines_long,
       aes(x = reorder(Substance, Concentration, FUN = median), y = Concentration
)) +
  geom_boxplot(color = "darkred") +
  labs(x = "Substance",
       y = "Concentration (%)") +
  theme_minimal()+
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_y_log10() +
  scale_x_discrete(labels = chem_labels)
a_log



```

## plot correlations between certain constituents

```{r correlation-plots}
# Correlation plots between constituents

# Select only the constituent columns (excluding metadata columns)
constituents <- dt_fines |> 
  select(SiO2, TiO2, Al2O3, Fe2O3, MnO, MgO, CaO, Na2O, K2O, P2O5, SO3) |> 
  drop_na()  # Remove rows with any missing values

# Calculate Spearman correlation matrix
cor_matrix <- cor(constituents, method = "spearman", use = "complete.obs")

# Calculate p-values for Spearman correlation using Hmisc
cor_result <- cor.mtest(as.matrix(constituents), type = "spearman")
p_matrix <- cor_result$p

# Create correlation plot
corrplot(cor_matrix, 
         method = "color", # Fill tiles with colors
         diag= F,        # Hide diagonal (self-correlations)
         type = "upper",              # Show both upper and lower triangle
         col = COL2('RdBu', 9),    # Standard blue to red color scheme
         tl.col = "black",           # Text label color: black
         addCoef.col = "black",      # Coefficient text color: black
         p.mat = p_matrix,           # P-value matrix for significance
         sig.level = 0.55,           # Alpha = 0.05
         insig = "label_sig",        # Add significance labels (* for p < 0.05)
         pch.cex = 0.9,              # Size of significance markers
         pch.col = "black",          # Color of significance markers
         number.cex = 1,             # Size of correlation coefficients
         tl.cex = 1.5,               # Size of text labels
         addCoefasPercent = FALSE  # Show coefficients as decimals, not percentages
         )               


#save to file (png, width=8in, height=8in, res=300)

ggsave("output/correlation_plot_fines.png", width = 8, height = 8, dpi = 300)

```


## add terniary diagrams 

in the followwing plot, the terniary diagram for CaO, MgO, P2O5 is shown; point color: deposit.

```{r ternary-diagram}
# Load Ternary package for ternary plots
if(!require(Ternary)) {
  install.packages("Ternary")
}
library(Ternary)

# Prepare data for ternary plot
# Ternary package requires a matrix with 3 columns
ternary_data <- dt_fines |> 
  select(CaO, MgO, P2O5, Deposit) |> 
  drop_na()

# Normalize data to specified limits: CaO max=100, MgO max=50, P2O5 max=10
# Convert to proportions that sum to 1
ternary_data_normalized <- ternary_data |>
  mutate(
    CaO_norm = CaO / 100,
    MgO_norm = MgO / 50,
    P2O5_norm = P2O5 / 10,
    total = CaO_norm + MgO_norm + P2O5_norm,
    CaO_tern = CaO_norm / total * 100,
    MgO_tern = MgO_norm / total * 100,
    P2O5_tern = P2O5_norm / total * 100
  )

# Create ternary plot
TernaryPlot(
  atip = "CaO (max=100%)",
  btip = "MgO (max=50%)",
  ctip = expression(paste(P[2]*O[5], " (max=10%)")),
  axis.labels = seq(0, 100, by = 10),
  grid.minor.lines = 0
)

# Add points colored by deposit
deposits <- unique(ternary_data_normalized$Deposit)
colors <- scales::hue_pal()(length(deposits))

for(i in seq_along(deposits)) {
  deposit_subset <- ternary_data_normalized |> filter(Deposit == deposits[i])
  TernaryPoints(
    deposit_subset[, c("CaO_tern", "MgO_tern", "P2O5_tern")],
    col = colors[i],
    pch = 16,
    cex = 1.2
  )
}

# Add legend
legend(
  "topright",
  legend = deposits,
  col = colors,
  pch = 16,
  bty = "n",
  title = "Deposit"
)

  

```



```{r}
# Install package plotly for interactive ternary plots
if(!require(plotly)) {
  install.packages("plotly")
}
library(plotly)

# Prepare ternary data (already defined earlier in the document)
ternary_data <- dt_fines |> 
  select(CaO, MgO, P2O5, Deposit) |> 
  drop_na()

# Normalize data to sum to 100 for ternary plot
# Each component is scaled by its maximum value, then normalized
ternary_data_normalized <- ternary_data |>
  mutate(
    CaO_scaled = CaO ,
    MgO_scaled = MgO / 30 * 100,  # Scale MgO to a max of 30
    P2O5_scaled = P2O5 / 5 * 100, # Scale P2O5 to a max of 10

    # Add formatted tooltip text
    tooltip_text = paste0(
      "Deposit: ", Deposit, "<br>",
      "CaO: ", round(CaO, 2), "%<br>",
      "CaO_scaled: ", round(CaO_scaled, 2), "%<br>",
      "MgO: ", round(MgO, 2), "%<br>",
      "MgO_scaled: ", round(MgO_scaled, 2), "%<br>",
      "P₂O₅: ", round(P2O5, 2), "%",
      "P₂O₅_scaled: ", round(P2O5_scaled, 2), "%"
    )
  )


# Create interactive ternary plot with custom axis ranges
fig <- ternary_data_normalized |> 
  plot_ly(
    type = 'scatterternary',
    mode = 'markers',
    a = ~CaO_scaled,
    b = ~MgO_scaled,
    c = ~P2O5_scaled,
    text = ~tooltip_text,
    hoverinfo = 'text',
    color = ~Deposit,
    colors = scales::hue_pal()(length(unique(ternary_data_normalized$Deposit))),
    marker = list(
      size = 10,
      line = list(width = 1)
    )
  ) |>
  layout(
    ternary = list(
      sum = 100,
      aaxis = list(
        title = 'CaO (%)',
        titlefont = list(size = 20),
        tickfont = list(size = 15),
        tickvals = seq(0, 100, 10),
        ticktext = seq(0, 100, 10)
      ),
      baxis = list(
        title = 'MgO (%)',
        titlefont = list(size = 20),
        tickfont = list(size = 15),
        tickvals = seq(0, 100, 10),
        ticktext = seq(0, 30, 3)
      ),
      caxis = list(
        title = 'P<sub>2</sub>O<sub>5</sub> (%)',
        titlefont = list(size = 20),
        tickfont = list(size = 15),
        tickvals = seq(0, 100, 10),
        ticktext = seq(0, 5, 0.5)
      )
    ),
    margin = list(l = 80, r = 80, t = 80, b = 80)
  )

fig

```